Two subtle problems with over-representation analysis

### 2024-07-10 manuscript/manuscript.Rmd
works
refine ms

The second problem is that adjusting p-values for multiple testing, also known as
false-discovery rate correction, is implemented improperly.
Determining which pathways should be subject to false discovery rate correction depends on
how a detection threshold is set.
Ideally, a pathway should be considered detected based on the presence of a predetermined
number of member genes in the background set.
However, some tools use the foreground set to define a pathway as detected.
This is problematic, as the act of trying to find an intersection between a foreground gene list
and a pathway is a test, even if there are no common genes.
This results in pathways being discarded from the analysis if no common genes are found in
the foreground, despite many genes being present in the background list.
This artificially reduces the number of tests conducted and makes FDR values appear lower
than they should, potentially raising the rate of false positives.
We expect this problem to have a more severe effect when the foreground list is small.

@@

![**Figure 1B. Impact of background problem is worse for smaller gene set libraries.** Mean
Jaccard similarity index was calculated for original and corrected ORA for seven independent
datasets. The number of genes represented in one or more functional sets is shown on the x-axis,
and the mean Jaccard index is shown on the y-axis.](../analysis/fig1_ORAjac.png "Figure 1B
shows that the background bug impacts smaller gene sets libraries like KEGG and HPO more
severely.")

@@

![**Figure 2B. Impact of FDR problem is worse for shorter foreground gene lists.** Mean Jaccard
index between original and corrected analysis were calculated for seven RNA-seq datasets and nine
gene set libraries. Only GO, Reactome, Cellmarkers and TFT GTRD are shown.](
../analysis/fig2_ORAsz.png "Figure 2B shows that the impact of the FDR bug is worse when
foreground gene lists are shorter than 500.")

@@

ClusterProfiler is not the only tool to suffer from these problems, they are widespread
(**Table 1**). Still, there are some tools free of such problems.
For example, fora is a good option for R-based workflows, however the function does not calculate the fold enrichment score,
so users will need to do it themselves [@Korotkevich2016-gd].
ShinyGO is an excellent alternative for web based analysis [@Ge2020-lu].
It doesn't suffer from the two issues described here, it is easy to use,
and has a high degree of reproducibility thanks to its stand-alone versioned docker
container option for local execution.
One problem that all of these tools listed in Table 1 have is that a background list is optional.
Sometimes this feature is hidden under "advanced options".
Given the strong biological and technical biases that are present in contemporary omics such as including single cell and spatial transcriptomics,
it should be mandatory step for users to provide a custom background list.

@@

The purpose of this work was to thoroughly characterise two subtle problems with ORA, but there are some
bigger problems noted.
Firstly, all variations of ORA were less accurate as compared to FCS in this simulation work, confirming a
previous report [@Kaspi2020-rn].
Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS
was relatively stable.
This means researchers can be more confident with FCS results, even if the data are
noisy.
Higher recall of FCS also means that researchers will have more power to identify pathways
relevant to the biological processes being studied.
Secondly, many ORA tools do not provide enrichment scores in their default output tables.
Enrichment scores are a surrogate for effect size, and without them, users may over-rely
on significance scores when interpreting enrichment results [@Ziemann2024-ts].
Lastly, ORA results can vary dramatically depending on the number of genes selected in the
foreground, a mostly arbitrary choice.

>>>>

The second problem is that adjusting p-values for multiple testing, also known as
false-discovery rate correction, is sometimes implemented improperly.
Determining which pathways should be subject to false discovery rate correction depends on
how a detection threshold is set.
Ideally, a pathway should be considered detected based on the presence of a predetermined
number of member genes in the background set.
However, some tools use the foreground set to define a pathway as detected.
This is problematic, as the act of trying to find an intersection between a foreground gene list
and a pathway is a test, even if there are no common genes.
This results in pathways being discarded from the analysis if no common genes are found in
the foreground, despite many genes being present in the background list.
This artificially reduces the number of tests conducted and makes FDR values appear lower
than they should, potentially raising the rate of false positives.
We expect this problem to have a more severe effect when the foreground list is small.

@@

![**Figure 1B. Impact of the background problem is worse for smaller gene set libraries.** Mean
Jaccard similarity index was calculated for original and corrected ORA for seven independent
datasets. The number of genes represented in one or more functional sets is shown on the x-axis,
and the mean Jaccard index is shown on the y-axis.](../analysis/fig1_ORAjac.png "Figure 1B
shows that the background bug impacts smaller gene sets libraries like KEGG and HPO more
severely.")

@@

![**Figure 2B. Impact of the FDR problem is worse for shorter foreground gene lists.** Mean Jaccard
index between original and corrected analysis were calculated for seven RNA-seq datasets and nine
gene set libraries. Only GO, Reactome, Cellmarkers and TFT GTRD are shown.](
../analysis/fig2_ORAsz.png "Figure 2B shows that the impact of the FDR bug is worse when
foreground gene lists are shorter than 500.")

@@

ClusterProfiler is not the only tool to suffer from these problems, they are widespread
(**Table 1**). Still, there are some tools free of such problems.
For example, fora is a good option for R-based workflows, however the function does not calculate the fold enrichment score,
so users will need to do it themselves [@Korotkevich2016-gd].
ShinyGO is an excellent alternative for web based analysis [@Ge2020-lu].
It doesn't suffer from the two issues described here, it is easy to use,
and has a high degree of reproducibility thanks to its stand-alone versioned docker
container option for local execution.
One problem that all of these tools listed in Table 1 have is that a background list is optional.
Sometimes this feature is hidden under "advanced options".
Given the strong biological and technical biases that are present in contemporary omics such as including single cell and spatial transcriptomics,
it should be a mandatory step for users to provide a custom background list.

@@

The purpose of this work was to thoroughly characterise two subtle problems with ORA, but there are some
bigger problems noted.
Firstly, all variations of ORA were less accurate as compared to FCS in this simulation work, confirming a
previous report [@Kaspi2020-rn].
Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS
was relatively stable.
This means researchers can be more confident with FCS results, even if the data are
noisy.
Higher recall of FCS also means that researchers will have more power to identify pathways
relevant to the biological processes being studied.
Secondly, many ORA tools do not provide enrichment scores in their default output tables.
Enrichment scores are a surrogate for effect size, and without them, users may over-rely
on significance scores when interpreting enrichment results.
Lastly, ORA results can vary dramatically depending on the number of genes selected in the
foreground, a mostly arbitrary choice.
